Kalman tracking and parameter estimation of continuous gravitational waves with a pulsar timing array

### 2023-11-04 docs/manuscript/paper.tex
all correctins up to appendix

\begin{figure*}
	\includegraphics[width=\textwidth, height =\textwidth ]{images/small_h_posterior_10}
	\caption{Posterior distribution of the GW source parameters $\boldsymbol{\theta}_{\rm gw}$ for the representative system in Table \ref{tab:parameters_and_priors}, for a single realisation of the system noise. The horizontal and vertical orange lines indicate the true injected values. The contours in the two-dimensional histograms mark the (0.5, 1, 1.5, 2)-$\sigma$ levels after marginalizing over all but two parameters. The one-dimensional histograms correspond to the joint posterior distribution marginalized over all but one parameter. The supertitles of the marginalized one-dimensional histograms specify the posterior median and the 0.16 and 0.84 quantiles. We plot the scaled variables $\Omega_{\rm nHz} = \Omega \times 10^{9}$ Hz and $h_{0, \times 10^{15}} = h_0 \times 10^{15}$. The Kalman filter and nested sampler estimate accurately all seven parameters in ${\boldsymbol{\theta}}_{\rm gw}$. The horizontal axes span a subset of the prior domain for all seven parameters.}
	\label{fig:corner_plot_1}
\end{figure*}

@@

\subsection{Multiple noise realisations: dispersion of outcomes} \label{sec:multiple_noise}
\begin{figure*}
	\includegraphics[width=\textwidth, height =\textwidth]{images/stacked_GW_plot_small_h_definitive3}
	\caption{Same as Figure \ref{fig:corner_plot_1} but for nine realisations of the noise processes, with curves coloured uniquely. The supertitles of the one-dimensional histograms record the posterior median and the 0.16 and 0.84 quantiles of the median realisation. The known, injected value lies within the 90\% credible interval for 58 out of the 63 combinations of seven parameters and nine noise realizations. There is an appreciable dispersion among the peaks of the one-dimensional posteriors, with coefficient of variation $\sim10 \%$ across the 63 combinations. A slight skew-bias is apparent in some of the parameters, e.g. $\iota$. Weak correlations can be seen between $\Omega$ and $\Phi_0$, $\psi$ and $\alpha$, and $\iota$ and $h_0$.} 
	\label{fig:corner_plot_2}
\end{figure*}
\begin{table}
	\centering
	\begin{tabular}{ccll}
		\toprule
		Parameter & Injected value & Units & $W_{1, \rm median}$  \\
		\hline
		$\Omega$     &   $5 \times 10^{-7}$ & Hz & $10^{-9}$ \\
		$\Phi_0$          & $0.20$ & rad & $0.13$ \\
		$\psi$              & $2.50$ & rad & $0.16$ \\
		$\iota$             & $1.0$ & rad & $0.14$ \\ 
		$\delta$              & $1.0$  & rad & $0.09$ \\
		$\alpha$          & $1.0$  & rad & $0.17$\\
		$h_0$            & $5 \times 10^{-15}$ & --- & $8 \times 10^{-16}$ \\
		\bottomrule
	\end{tabular}
	\caption{Median value of the Wasserstein distance $W_{1, \rm median}$, for each parameter in $\boldsymbol{\theta}_{\rm gw}$, calculated across the $10^3 \choose 2$ pairs of probability posteriors, for the $10^3$ noise realisations in Figure \ref{fig:pairwise_wasserstein}. $W_{1, \rm median}$ is generally smaller than the prior domain.}
	\label{tab:Wasserstein}
\end{table}

@@

Table \ref{tab:Wasserstein} summarizes the WD between the $5\times 10^5$ pairs of one-dimensional posteriors across the $10^3$ realizations, for each of the seven parameters in ${\boldsymbol{\theta}}_{\rm gw}$. The median $W_1$ for each parameter is tabulated in the rightmost column and denoted by $W_{1,{\rm median}}$. As a percentage of the known, injected value, $W_{\rm 1,median}$ ranges from  0.2\% for $\Omega$ to 67\% for $\Phi_0$. These values are an approximate measure of the natural dispersion in parameter estimates when the analysis scheme is applied to real, astronomical data, where the true parameter values are unknown. Moreover, as a percentage of the prior domain, $W_{\rm 1,median}$ ranges from $8.1 \times 10^{-4}$ \% for $h_0$ to 2.7\% for $\psi$. These values confirm that the nested sampler converges reliably to a single, narrow peak without railing against the prior bounds for a wide variety of noise realizations. Further analysis of the WD is performed in Appendix \ref{sec:wasserstein}.

\subsection{Detectability versus $h_0$} \label{sec:detection}
We frame the problem of detecting a GW in noisy PTA data in terms of the Bayesian model selection procedure described in Section \ref{sec:model_selection}. In Equation \eqref{eq:bayes} $\mathcal{M}_1$ is the Earth-terms-only model with a GW present, i.e. the state-space model with a Kalman filter based on Equation \eqref{eq:measuremen_earth}. The Bayes factor, $\beta$, as defined in Equation \eqref{eq:bayes} is presented in Figure \ref{fig:bayes} for the representative source in Table \ref{tab:parameters_and_priors}, except that we now vary the source amplitude, $h_0$, from $10^{-15}$ (undetectable) to $10^{-12}$ (easily detectable). To control the test, the noise processes in the synthetic data are identical realisations for each value of $h_0$; the only change from one $h_0$ value to the next is $h_0$ itself. \footnote{Changing the noise realizations as well, from one value of $h_0$ to the next, adds uninformative scatter to the trend in Figure \ref{fig:bayes}.} \newline

@@

We analyse 200 injections constructed by fixing $h_0 = 5 \times 10^{-15}$ and $\iota =1.0$ and drawing the remaining five elements of $\boldsymbol{\theta}_{\rm gw}$ randomly from the prior distributions defined in Table \ref{tab:parameters_and_priors}. We fix $h_0$ and $\iota$ in order to maintain an approximately constant signal-to-noise ratio (SNR) across the 200 injections. For each injection we compute the posterior distribution of ${\boldsymbol{\theta}}_{\rm gw}$. To summarise the results we use a parameter-parameter (PP) plot \citep{doi:10.1198/106186006X136976}. A PP plot displays the fraction of injections included within a given credible interval of the estimated posterior, plotted as a function of the credible interval itself. In the ideal case of perfect recovery, the PP plot should be a diagonal line of unit slope. \newline

@@

\subsection{Identifiability}\label{sec:identif}
\begin{figure}
	\centering
	\includegraphics[width=\columnwidth]{images/stacked_GW_plot_iota_h}
	\caption{Corner plot of the joint posterior distribution of the GW source parameters $h_0$ and $\iota$ for the representative system in Table 1 in the high-SNR regime ($h_0 = 10^{-12}$). Nine coloured curves are plotted for nine random noise realizations. The posteriors overlap less than for the other five parameters in ${\boldsymbol{\theta}}_{\rm gw}$. A correlation between $h_0$ and $\iota$ is evident in the bottom left panel.}
	\label{fig:just_iota_h}
\end{figure}
\begin{figure}
	\centering
	\includegraphics[width=\columnwidth]{images/likelihood_surface}
	\caption{Log-likelihood $\log \mathcal{L}$ (Equation \ref{eq:likelihood}) contours in the $\iota$- $h_0$ plane for a single noise realisation. The red dot marks the maximum of $\log {\cal L}$. The colour scale is normalised with respect to the absolute value of the maximum. An S-shaped likelihood ridge is visible, which makes it difficult for the nested sampler to converge to a unique value.}
	\label{fig:likelihood_surface}
\end{figure}
In Figure \ref{fig:just_iota_h} we plot the one- and two-dimensional posteriors for $\iota$ and $h_0$ for the high-SNR representative system for nine noise realisations. The figure is entirely analogous to Figure \ref{fig:corner_plot_2}, but the corner plot displays $\iota$ and $h_0$ rather than all elements of $\boldsymbol{\theta}_{\rm gw}$. One observes four things: (i) the plotted posteriors are unimodal but offset systematically from the true, injected values, due to the bias discussed in Section 7.2; (ii) a correlation between $h_0$ and $\iota$ is visible in the two-dimensional contours; (iii) the posteriors do not overlap tightly, a trend seen also in the $\delta$ posterior in Figure \ref{fig:corner_plot_2}; and (iv) features (i)--(iii) occur, even though the SNR is high. One expects a priori that the parameter estimates should become more accurate, as the SNR increases, but this is not true for $h_0$ and $\iota$. Specifically, across $10^3$ noise realisations, the median WD for $\iota$ is $W_{1, \rm median}=0.24$ rad and for $h_0$ is  $W_{1, \rm median}=1.6 \times 10^{-13}$. Both $W_{1,{\rm median}}$ values are higher (as a fraction of the true, injected value) than in the lower-SNR tests conducted in Sections \ref{sec:rep_example} and \ref{sec:parameter_space}. In contrast, $W_{1,\rm median}$ for the other five parameters in $\boldsymbol{\theta}_{\rm gw}$ decreases, as the SNR increases. \newline

We infer different $h_0$ and $\iota$ posteriors for different noise realizations, because $h_0$ and $\iota$ are weakly identifiable. Identifiability is a formal concept from the theory of signal processing, which is applied routinely to engineering problems \citep{e5be7c83a0d24500826f6e1b414d1733}. It refers to whether it is theoretically possible to infer unique and accurate parameter values, given the measured data and the model structure. Sources of non-identifiability are generally categorised as either ``structural'' , arising from the structure of the model (e.g. its invertibility), or ``practical'', arising from insufficient data or measurement errors \citep{GUILLAUME2019418}. Regarding $\iota$ and $h_0$, the source is structural. Equations \eqref{eq:hphx} and \eqref{eq:hphx2} contain a weak degeneracy between $\iota$ and $h_0$. For example, the cross polarisation strain $h_{\times}$ involves a trade-off between $\iota$ and $h_0$, which is broken by knowing $h_+$ via Equation \eqref{eq:hphx}. However, the measured strain is a linear combination of $h_\times$ and $h_+$; the two polarizations are not measured independently, which makes it challenging to separate $\iota$ and $h_0$. In fact $\iota$ and $h_0$ are structurally identifiable --- but only weakly. That is, the $\iota$-$h_0$ likelihood contours plateau to a ridge with a small, non-zero gradient close to the maximum likelihood solution. The likelihood ridge is presented in Figure \ref{fig:likelihood_surface} for a particular noise realisation. Whilst a single, unique maximum does exist (labelled by the red dot in the figure), it lies along an S-shaped ridge where the likelihood values are all similar. Consequently it is difficult for the nested sampling algorithm to settle on the red dot; it trades off accuracy in one parameter against the other, depending on the particular realisation of the noise. By way of verification, if we fix one parameter at its true value (i.e. set a delta function prior) during the inference analysis, the posteriors for the other parameter overlap indistinguishably  for different noise realisations, i.e. $W_1 \to 0$.

@@

\begin{figure}
	\centering
	\includegraphics[width=\columnwidth]{images/likelihood_iota_new}
	\caption{Log-likelihood returned by the Kalman filter as a function of inclination $\iota$, using the Earth terms in the Kalman filter (Equation \ref{eq:measuremen_earth}, blue line), the pulsar terms in the Kalman filter (Equation \ref{eq:measurement}, orange line), and a control experiment that drops the pulsar terms from both the synthetic data and the Kalman filter (green line). All other parameters are fixed at their true injected value from Table \ref{tab:parameters_and_priors}. The coloured dashed vertical lines mark the maximum for the corresponding model; the green and orange dashed lines overlap at this resolution. For the Earth-terms model the likelihood maximum ($\iota \approx 0.7$) does not coincide with the injected value ($\iota = 1.0$). For the pulsar-terms model the maximum and injected value do coincide. 10 noise realisations are plotted for each situation but overlap at this resolution; they are indistinguishable to the eye.}
	\label{fig:likelihood_surface_iota}
\end{figure}

Sections \ref{sec:rep_example} and \ref{sec:parameter_space} present preliminary evidence that the inferred value of ${\boldsymbol{\theta}}_{\rm gw}$ (e.g. the medians of the one-dimensional marginalised posteriors) is biased away from the true injected value. Figure \ref{fig:bias_for_large_h} displays a corner plot for $\boldsymbol{\theta}_{\rm gw}$, analogous to Figure \ref{fig:corner_plot_1}, but for high SNR $(h_0 = 10^{-12})$. Except for the elements $\Omega$ and $h_0$, the injected value lies outside the 90\% credible interval, to the extent that it falls outside the plotted domain. The deviation is most severe for $\iota$ with a bias of $\sim 0.3$ rad but is also present to a lesser extent in other static parameters such as $\delta,\alpha$ and $\psi$. Similar results are obtained for other noise realisations. Whilst $h_0$ does not exhibit any noticeable bias for the particular noise realisation generating Figure \ref{fig:bias_for_large_h}, in general it does exhibit a correlated bias with $\iota$ across multiple noise realisations, as noted in the bottom left panel of Figure \ref{fig:just_iota_h}. In this section we demonstrate how the bias stems from dropping the pulsar terms described in Section \ref{sec:parameter_estim}. \newline

In Figure \ref{fig:likelihood_surface_iota} we plot the log-likelihood returned by the Kalman filter, viz. Equation \eqref{eq:likelihood}, as a function of $\iota$ (i.e. $\log \mathcal{L}(\iota)$) across the prior domain, holding all other parameters constant at the true injected value of the representative system in Table \ref{tab:parameters_and_priors}. We generate 10 plots of $\log \mathcal{L}(\iota)$ corresponding to 10 separate realisations of the system noise. The solid blue curves only includes the Earth terms in the measurement equation (i.e. using Equation \eqref{eq:measuremen_earth} in the Kalman filter). The solid orange curves include the Earth and pulsar terms in the measurement equation (i.e. using Equation \eqref{eq:measurement}). The solid green curve corresponds to a control experiment, where we drop the pulsar terms from both the synthetic data and the measurement equation. The vertical dashed lines mark the corresponding maxima, which nested sampling seeks. \newline

The key observation from Figure \ref{fig:likelihood_surface_iota} is that the location of the maximum ($\iota \approx 0.7 \, {\rm rad}$) does not coincide with the true, injected value ($\iota = 1.0 \, {\rm rad}$), when only the Earth terms are included. This is the origin of the bias in the one-dimensional marginalised posteriors in Figure \ref{fig:bias_for_large_h}. In contrast, the maximum aligns with the true, injected value, when the pulsar terms are included. As a cross-check, agreement is also achieved, if we drop the pulsar terms from the synthetic data (which is not the situation in reality, of course) and use the Earth terms only for inference. Similar results are obtained for other parameters, e.g. $\delta$. A discussion on why the bias is present in particular elements of $\boldsymbol{\theta}_{\rm gw}$, and the implications of the bias for continuous wave searches, are presented in Appendix \ref{sec:bias_appendix}. Dropping the pulsar terms biases several of the inferred parameters (e.g.\ those that affect the GW amplitude), not just the sky position  \citep[e.g.][]{Zhupulsarterms,Chen2022}. The bias is not particular to our method; it is shared by all likelihood-based methods that do not include the pulsar terms in the inference model. Generalizing the Kalman tracker and nested sampler to correct for the bias will be a key goal of a forthcoming paper.

@@

In this paper we demonstrate a new method for the detection and parameter estimation of GWs from individual, monochromatic SMBHBs in PTA data. The new method is complementary to traditional approaches. We track the evolution of the intrinsic pulsar timing noise explicitly via a state-space method, rather than subtracting the ensemble-averaged statistics. That is, we disentangle statistically the specific time-ordered realisation of the timing noise from the GW-induced modulations, and thereby infer the GW source parameters conditional on the specific observed realisation of the noisy data. We implement a Kalman filter in order to track the intrinsic rotational state evolution of the pulsar and combine it with a Bayesian sampling framework to estimate the posterior distributions of each static parameter, as well as the associated Bayesian evidence (marginalised likelihood) of the model. \newline

 We test the new method on synthetic data and find that it detects injected signals successfully and estimates their static parameters accurately with relatively low computational cost. We initially focus on a single, astrophysically representative, SMBHB GW source observed synthetically by the 12.5-year NANOGrav pulsars with $T_{\rm obs} = 10 \, {\rm years}$. The minimum detectable strain is estimated to be $\min(h_0) \approx 4 \times 10^{-15}$ for $\iota=1.0$ rad. We then repeat the parameter estimation exercise for 1000 noise realisations to compute the natural dispersion (``cosmic variance'') in the recovered values and hence quantify the statistical accuracy of the method in a real astrophysical application, when the true parameter values are unknown. Consistent posteriors are obtained for most realisations, with the median Wasserstein distance limited to $\lesssim 3\%$ of the width of the prior domain (i.e.\ negligible railing). The median Wasserstein distance as a fraction of the injected value, an approximate measure of the natural dispersion, ranges from  0.2\% for $\Omega$ to 67\% for $\Phi_0$. \newline

Exploration of a broader SMBHB parameter domain at fixed $\iota$ and $h_0$ via 200 randomly sampled parameter vectors suggests at the presence of a bias in the estimates of the static parameters $\delta$ and $\alpha$, (and to a lesser extent in $\psi$ and $\Phi_0$). The bias is examined for a specific SMBHB GW source in the limit of high-SNR. It is greatest for $\iota$ at $\sim 0.3$ rad, but smaller biases of $ \lessapprox 0.1$ rad are also observed in $\Phi_0, \psi, \delta$ and $\alpha$. The bias is shown to result from dropping the pulsar terms from the measurement equation in the Kalman filter. The method also exhibits a trade-off between $h_0$ and $\iota$, which follows formally from the weak identifiability of $h_0$ and $\iota$. \newline

>>>>

\begin{figure*}
	\includegraphics[width=\textwidth, height =\textwidth ]{images/small_h_posterior_10}
	\caption{Posterior distribution of the GW source parameters $\boldsymbol{\theta}_{\rm gw}$ for the representative system in Table \ref{tab:parameters_and_priors}, for a single realisation of the system noise. The horizontal and vertical orange lines indicate the true injected values. The contours in the two-dimensional histograms mark the (0.5, 1, 1.5, 2)-$\sigma$ levels after marginalizing over all but two parameters. The one-dimensional histograms correspond to the joint posterior distribution marginalized over all but one parameter. The supertitles of the marginalized one-dimensional histograms specify the posterior median and the 0.16 and 0.84 quantiles. We plot the scaled variables $10^9 \Omega$ (units: ${\rm rad \, s^{-1}}$) and $10^{15} h_0$. The Kalman filter and nested sampler estimate accurately all seven parameters in ${\boldsymbol{\theta}}_{\rm gw}$. The horizontal axes span a subset of the prior domain for all seven parameters.}
	\label{fig:corner_plot_1}
\end{figure*}

@@

\subsection{Multiple noise realisations: dispersion of outcomes} \label{sec:multiple_noise}
\begin{figure*}
	\includegraphics[width=\textwidth, height =\textwidth]{images/stacked_GW_plot_small_h_definitive2}
	\caption{Same as Figure \ref{fig:corner_plot_1} but for nine realisations of the noise processes, with curves coloured uniquely. The supertitles of the one-dimensional histograms record the posterior median and the 0.16 and 0.84 quantiles of the median realisation. The known, injected value lies within the 90\% credible interval for 58 out of the 63 combinations of seven parameters and nine noise realizations. There is an appreciable dispersion among the peaks of the one-dimensional posteriors, with coefficient of variation $\sim10 \%$ across the 63 combinations. A slight skew-bias is apparent in some of the parameters, e.g. $\iota$. Weak correlations can be seen between $\Omega$ and $\Phi_0$, $\psi$ and $\alpha$, and $\iota$ and $h_0$.} 
	\label{fig:corner_plot_2}
\end{figure*}
\begin{table}
	\centering
	\begin{tabular}{ccll}
		\toprule
		Parameter & Injected value & Units & $W_{1, \rm median}$  \\
		\hline
		$\Omega$     &   $5 \times 10^{-7}$ & Hz & $10^{-9}$ \\
		$\Phi_0$          & $0.20$ & rad & $0.13$ \\
		$\psi$              & $2.50$ & rad & $0.16$ \\
		$\iota$             & $1.0$ & rad & $0.14$ \\ 
		$\delta$              & $1.0$  & rad & $0.09$ \\
		$\alpha$          & $1.0$  & rad & $0.17$\\
		$h_0$            & $5 \times 10^{-15}$ & --- & $8 \times 10^{-16}$ \\
		\bottomrule
	\end{tabular}
	\caption{Median value of the Wasserstein distance $W_{1, \rm median}$, for each parameter in $\boldsymbol{\theta}_{\rm gw}$, calculated across the $10^3 \choose 2$ pairs of probability posteriors, for the $10^3$ noise realisations in Figure \ref{fig:pairwise_wasserstein}. $W_{1, \rm median}$ is generally smaller than the prior domain.}
	\label{tab:Wasserstein}
\end{table}

@@

Table \ref{tab:Wasserstein} summarizes the WD between the $5\times 10^5$ pairs of one-dimensional posteriors across the $10^3$ realizations, for each of the seven parameters in ${\boldsymbol{\theta}}_{\rm gw}$. The median $W_1$ for each parameter is tabulated in the rightmost column and denoted by $W_{1,{\rm median}}$. As a percentage of the known, injected value, $W_{\rm 1,median}$ ranges from  0.2\% for $\Omega$ to 67\% for $\Phi_0$. These values quantify approximately the natural dispersion in parameter estimates when the analysis scheme is applied to real, astronomical data, where the true parameter values are unknown. Moreover, as a percentage of the prior domain, $W_{\rm 1,median}$ ranges from $8.1 \times 10^{-4}$ \% for $h_0$ to 2.7\% for $\psi$. These values confirm that the nested sampler converges reliably to a single, narrow peak without railing against the prior bounds for $10^3$ noise realizations. Further analysis of the WD is performed in Appendix \ref{sec:wasserstein}.

\subsection{Detectability versus $h_0$} \label{sec:detection}
We frame the problem of detecting a GW in noisy PTA data in terms of the Bayesian model selection procedure described in Section \ref{sec:model_selection}. In Equation \eqref{eq:bayes} $\mathcal{M}_1$ is the Earth-term-only model with a GW present, i.e. the state-space model with a Kalman filter based on Equation \eqref{eq:measuremen_earth}. The Bayes factor, $\beta$, as defined in Equation \eqref{eq:bayes} is presented in Figure \ref{fig:bayes} for the representative source in Table \ref{tab:parameters_and_priors}, except that we now vary the source amplitude, $h_0$, from $10^{-15}$ (undetectable) to $10^{-12}$ (easily detectable). To control the test, the noise processes in the synthetic data are identical realisations for each value of $h_0$; the only change from one $h_0$ value to the next is $h_0$ itself. \footnote{Changing the noise realizations as well, from one value of $h_0$ to the next, adds uninformative scatter to the trend in Figure \ref{fig:bayes}.} \newline

@@

We analyse 200 injections constructed by fixing $h_0 = 5 \times 10^{-15}$ and $\iota =1.0$ rad and drawing the remaining five elements of $\boldsymbol{\theta}_{\rm gw}$ randomly from the prior distributions defined in Table \ref{tab:parameters_and_priors}. We fix $h_0$ and $\iota$ in order to maintain an approximately constant signal-to-noise ratio (SNR) across the 200 injections. For each injection we compute the posterior distribution of ${\boldsymbol{\theta}}_{\rm gw}$. To summarise the results we use a parameter-parameter (PP) plot \citep{doi:10.1198/106186006X136976}. A PP plot displays the fraction of injections included within a given credible interval of the estimated posterior, plotted as a function of the credible interval itself. In the ideal case of perfect recovery, the PP plot should be a diagonal line of unit slope. \newline

@@

\subsection{Identifiability}\label{sec:identif}
\begin{figure}
	\centering
	\includegraphics[width=\columnwidth]{images/stacked_GW_plot_iota_h}
	\caption{Corner plot of the joint posterior distribution of the GW source parameters $h_0$ and $\iota$ for the representative system in Table 1 in the high-SNR regime ($h_0 = 10^{-12}$). Nine coloured curves are plotted for nine random noise realizations. The posteriors overlap less than for the other five parameters in ${\boldsymbol{\theta}}_{\rm gw}$. A correlation between $h_0$ and $\iota$ is evident in the bottom left panel.}
	\label{fig:just_iota_h}
\end{figure}
\begin{figure}
	\centering
	\includegraphics[width=\columnwidth]{images/likelihood_surface}
	\caption{Log-likelihood $\log \mathcal{L}$ (Equation \ref{eq:likelihood}) contours in the $\iota$-$h_0$ plane for a single noise realisation. The red dot marks the maximum of $\log {\cal L}$. The colour scale is normalised with respect to the absolute value of the maximum. An S-shaped, flat-topped likelihood ridge is visible, which makes it difficult for the nested sampler to converge to a unique value.}
	\label{fig:likelihood_surface}
\end{figure}
In Figure \ref{fig:just_iota_h} we plot the one- and two-dimensional posteriors for $\iota$ and $h_0$ for the high-SNR representative system for nine noise realisations. The figure is entirely analogous to Figure \ref{fig:corner_plot_2}, but the corner plot displays $\iota$ and $h_0$ rather than all elements of $\boldsymbol{\theta}_{\rm gw}$. One observes four things: (i) the plotted posteriors are unimodal but offset systematically from the true, injected values, due to the bias discussed in Section 7.2; (ii) a correlation between $h_0$ and $\iota$ is visible in the two-dimensional contours; (iii) the posteriors do not overlap tightly, a trend seen also in the $\delta$ posterior in Figure \ref{fig:corner_plot_2}; and (iv) features (i)--(iii) occur, even though the SNR is high. One expects a priori that the parameter estimates should become more accurate, as the SNR increases, but this is not true for $h_0$ and $\iota$. Specifically, across $10^3$ noise realisations, the median WD for $\iota$ is $W_{1, \rm median}=0.24$ rad and for $h_0$ is  $W_{1, \rm median}=1.6 \times 10^{-13}$. Both $W_{1,{\rm median}}$ values are higher (as a fraction of the true, injected value) than in the lower-SNR tests conducted in Sections \ref{sec:rep_example} and \ref{sec:parameter_space}. In contrast, $W_{1,\rm median}$ for the other five parameters in $\boldsymbol{\theta}_{\rm gw}$ decreases, as the SNR increases. \newline

We infer different $h_0$ and $\iota$ posteriors for different noise realizations, because $h_0$ and $\iota$ are weakly identifiable. Identifiability is a formal concept from the theory of signal processing, which is applied routinely to engineering problems \citep{e5be7c83a0d24500826f6e1b414d1733}. It refers to whether it is theoretically possible to infer unique and accurate parameter values, given the measured data and the model structure. Sources of non-identifiability are generally categorised as either ``structural'' , arising from the structure of the model (e.g. its invertibility), or ``practical'', arising from insufficient data or measurement errors \citep{GUILLAUME2019418}. Regarding $\iota$ and $h_0$, the source is structural. Equations \eqref{eq:hphx} and \eqref{eq:hphx2} contain a weak degeneracy between $\iota$ and $h_0$. For example, the cross polarisation strain $h_{\times}$ involves a trade-off between $\iota$ and $h_0$, which is resolved by knowing $h_+$ via Equation \eqref{eq:hphx}. However, the measured strain is a linear combination of $h_\times$ and $h_+$; the two polarizations are not measured independently, which makes it challenging to separate $\iota$ and $h_0$. In fact $\iota$ and $h_0$ are structurally identifiable --- but only weakly. That is, the $\iota$-$h_0$ likelihood contours plateau to a flat-topped ridge with a small, non-zero gradient close to the maximum likelihood solution. The likelihood ridge is presented in Figure \ref{fig:likelihood_surface} for a particular noise realisation. Whilst a single, unique maximum does exist (labelled by the red dot in the figure), it lies along an S-shaped ridge where the likelihood values are all similar. Consequently it is difficult for the nested sampling algorithm to settle on the red dot; it trades off accuracy in one parameter against the other, depending on the particular realisation of the noise. By way of verification, if we fix one parameter at its true value (i.e. set a delta function prior) during the inference analysis, the posteriors for the other parameter overlap indistinguishably  for different noise realisations, i.e. $W_1 \to 0$.

@@

\begin{figure}
	\centering
	\includegraphics[width=\columnwidth]{images/likelihood_iota_new}
	\caption{Log-likelihood returned by the Kalman filter as a function of inclination $\iota$, using the Earth term in the Kalman filter (Equation \ref{eq:measuremen_earth}, blue curve), the pulsar terms in the Kalman filter (Equation \ref{eq:measurement}, orange curve), and a control experiment that drops the pulsar terms from both the synthetic data and the Kalman filter (green curve). All other parameters are fixed at their true injected value from Table \ref{tab:parameters_and_priors}. The coloured dashed vertical lines mark the maximum for the corresponding model; the green and orange dashed lines overlap at this resolution. For the Earth-term model the likelihood maximum ($\iota \approx 0.7$ rad) does not coincide with the injected value ($\iota = 1.0$ rad). For the pulsar-terms model the maximum and injected value do coincide. 10 noise realisations are plotted for each situation but overlap at this resolution; they are indistinguishable to the eye.}
	\label{fig:likelihood_surface_iota}
\end{figure}

Sections \ref{sec:rep_example} and \ref{sec:parameter_space} present preliminary evidence that the inferred value of ${\boldsymbol{\theta}}_{\rm gw}$ (e.g. the medians of the one-dimensional marginalised posteriors) is biased away from the true injected value. Figure \ref{fig:bias_for_large_h} displays a corner plot for $\boldsymbol{\theta}_{\rm gw}$, analogous to Figure \ref{fig:corner_plot_1}, but for high SNR $(h_0 = 10^{-12})$. Except for the elements $\Omega$ and $h_0$, the injected values lie outside the 90\% credible interval, and indeed fall outside the plotted domain. The deviation is most severe for $\iota$ with a bias of $\approx 0.3$ rad but is also present to a lesser extent in other static parameters such as $\delta,\alpha$ and $\psi$. Similar results are obtained for other noise realisations. Whilst $h_0$ does not exhibit any noticeable bias for the particular noise realisation generating Figure \ref{fig:bias_for_large_h}, in general it does exhibit a correlated bias with $\iota$ across multiple noise realisations, as noted in the bottom left panel of Figure \ref{fig:just_iota_h}. In this section we demonstrate how the bias stems from dropping the pulsar terms described in Section \ref{sec:parameter_estim}. \newline

In Figure \ref{fig:likelihood_surface_iota} we plot the log-likelihood returned by the Kalman filter, viz. Equation \eqref{eq:likelihood}, as a function of $\iota$ (i.e. $\log \mathcal{L}(\iota)$) across the prior domain, holding all other parameters constant at the true injected value of the representative system in Table \ref{tab:parameters_and_priors}. We generate 10 plots of $\log \mathcal{L}(\iota)$ corresponding to 10 separate realisations of the system noise. The solid blue curves only includes the Earth term in the measurement equation (i.e. using Equation \eqref{eq:measuremen_earth} in the Kalman filter). The solid orange curves include the Earth and pulsar terms in the measurement equation (i.e. using Equation \eqref{eq:measurement}). The solid green curve corresponds to a control experiment, where we drop the pulsar terms from both the synthetic data and the measurement equation. The vertical dashed lines mark the corresponding maxima, which nested sampling seeks. \newline

The key observation from Figure \ref{fig:likelihood_surface_iota} is that the location of the maximum ($\iota \approx 0.7 \, {\rm rad}$) does not coincide with the true, injected value ($\iota = 1.0 \, {\rm rad}$), when only the Earth terms are included. This is the origin of the bias in the one-dimensional marginalised posteriors in Figure \ref{fig:bias_for_large_h}. In contrast, the maximum aligns with the true, injected value, when the pulsar terms are included. As a cross-check, agreement is also achieved, if we drop the pulsar terms from the synthetic data (which is not the situation in reality, of course) and use the Earth term only for inference. Similar results are obtained for other parameters, e.g. $\delta$. A discussion about why the bias is present in particular elements of $\boldsymbol{\theta}_{\rm gw}$, and the implications of the bias for continuous wave searches, are presented in Appendix \ref{sec:bias_appendix}. Dropping the pulsar terms biases several of the inferred parameters (e.g.\ those that affect the GW amplitude), not just the sky position  \citep[e.g.][]{Zhupulsarterms,Chen2022}. The bias is not particular to our method; it is shared by all likelihood-based methods that do not include the pulsar terms in the inference model. Generalizing the Kalman tracker and nested sampler to correct for the bias will be a key goal of a forthcoming paper.

@@

In this paper we demonstrate a new method for the detection and parameter estimation of GWs from individual, monochromatic SMBHBs in PTA data. The new method is complementary to traditional approaches. We track the evolution of the intrinsic pulsar timing noise explicitly via a state-space method, rather than fitting for the ensemble-averaged power spectral density of the noise. That is, we disentangle statistically the specific time-ordered realisation of the timing noise from the GW-induced modulations, and thereby infer the GW source parameters conditional on the specific observed realisation of the noisy data. We implement a Kalman filter in order to track the intrinsic rotational state evolution of the pulsar and combine it with a Bayesian sampling framework to estimate the posterior distributions of each static parameter, as well as the associated Bayesian evidence (marginalised likelihood) of the model with and without a GW. \newline

 We test the new method on synthetic data and find that it detects injected signals successfully and estimates their static parameters accurately with relatively low computational cost. We initially focus on a single, astrophysically representative, SMBHB GW source observed synthetically by the 12.5-year NANOGrav pulsars with $T_{\rm obs} = 10 \, {\rm years}$. The minimum detectable strain is estimated to be $\min(h_0) \approx 4 \times 10^{-15}$ for $\iota=1.0$ rad. We then repeat the parameter estimation exercise for 1000 noise realisations to compute the natural dispersion (``cosmic variance'') in the recovered values and hence quantify the statistical accuracy of the method in a real astrophysical application, when the true parameter values are unknown. Consistent posteriors are obtained for most realisations. The median WD is limited to $\lesssim 3\%$ of the width of the prior domain (i.e.\ negligible railing). The median WD divided by the injected value, an approximate measure of the natural dispersion, ranges from  0.2\% for $\Omega$ to 67\% for $\Phi_0$. \newline

Exploration of a broader SMBHB parameter domain at fixed $\iota$ and $h_0$ via 200 randomly sampled parameter vectors reveals a bias in the estimates of the static parameters $\delta$ and $\alpha$, and to a lesser extent $\psi$ and $\Phi_0$. The bias is examined for a specific SMBHB GW source in the limit of high-SNR. It is greatest for $\iota$, amounting to $\approx 0.3$ rad.  Smaller biases of $\lesssim 0.1$ rad are also observed in $\Phi_0, \psi, \delta$ and $\alpha$. The bias is shown to result from dropping the pulsar terms from the measurement equation in the Kalman filter, consistent with the work of previous authors \citep{Zhupulsarterms,Chen2022}. The method also exhibits a trade-off between $h_0$ and $\iota$, which follows formally from the weak identifiability of $h_0$ and $\iota$. \newline

We emphasize that the Kalman tracker and nested sampler in this paper do not supplant traditional PTA analysis approaches; they complement traditional approaches and are most powerful when used in tandem. Relatedly, it is misleading to ask whether the Kalman tracker and nested sampler are more or less sensitive than traditional approaches for two reasons. First, one must still generalize the Kalman tracker and nested sampler to ingest TOAs directly, as happens traditionally, instead of ingesting a frequency time series, as in this paper. Generalizing the algorithm is a subtle task and is postponed to a forthcoming paper. Second, the Kalman tracker and nested sampler are conditional on a noise model that is related but different to the noise model in traditional approaches. The analysis in this paper assumes a specific, time-ordered realization of a mean-reverting Ornstein-Uhlenbeck process satisfying Equations \eqref{eq:spinevol}--\eqref{eq:xieqn}, whereas traditional analyses assume a stationary Gaussian process described by an ensemble-averaged, power-law power spectral density, whose amplitude and exponent are adjustable. Hence the sensitivities cannot be compared directly. Clarifying the similarities and differences between various approaches promises to be a fruitful avenue of future work. It is also a subject of attention in audio-band GW data analysis involving hidden Markov models applied to data from terrestrial long-baseline interferometers \citep{PhysRevD.102.023006,PhysRevD.105.022002,Abbott_2022SCO,2022PhRvD.106f2002A}. \newline
